---
title: "Final Project"
author: "Helen Liang"
date: "2023-12-20"
output: pdf_document
---

# Introduction

The COVID-19 pandemic has presented unprecedented challenges to public health, economies, and societies worldwide. Understanding the severity and spread of the virus, particularly in hard-hit areas such as the United States and New York, is crucial for policymakers, healthcare professionals, and researchers. It informs strategies for containment, mitigation, and preparedness for future public health emergencies.

To address the problem, I will utilize month-to-month % changes in COVID-19 cases and deaths for the US and New York. Furthermore, the data analysis will involve data Validation to ensure accuracy and consistency in the reported figures, comparative analysis to assess month-over-month changes to highlight trends and spikes, and contextual consideration to acknowledge external factors that can influence the data, such as testing rates and policy changes.

Analytic technique of the analysis will include time-series analysis to understand the dynamics over time and identify patterns and descriptive statistics to summarize the central tendencies, dispersion, and shape of the distributions of percentage changes. 

This analysis will provide stakeholders with a clear picture of the pandemic's progression, support evidence-based decisions, and contribute to preparedness for future health crises. The goal is to convert data into insights that will clarify the pandemic's path and consequences.

```{r, include = FALSE}
# set working directory
getwd()
setwd("/Users/helenliang/Library/Mobile Documents/com~apple~CloudDocs/NYU/Fall 2023/GPH-GU-2184/Final Project")
```

# Packages Required

The following packages are required to be installed and loaded prior to data preparation and data analysis to reproduce the codes and results throughout this project.

```{r, message = FALSE}
## Load Required Packages ##
library(readr) # reading csv data
library(tidyr) # data cleaning
library(dplyr) # data manipulation
library(knitr) # generating tables
library(lubridate) # handling date and time
library(ggplot2) # data visualization
```

# Data Preparation

This project involves the utilization of `us.csv` and `us-states.csv` datasets. Prior to assessing how these datasets have behaved over the course of 2020, we must acquire and clean the data. 

## Loading Data

The `us.csv` and `us-states.csv` datasets are originally sourced from The New York Times GitHub repository: [Coronavirus (Covid-19) Data in the United States](https://github.com/nytimes/covid-19-data), where it tracks daily U.S. COVID-19 cases since January 21, 2020, the beginning of the pandemic. 

These datasets contain COVID-19 data until January 13, 2021 and are accessible and downloadable through the `GPH-GU-2184 Intermediate Statistical Programming in R` course page on NYU Brightspace.

```{r, warning = FALSE}
# load us.csv data
us <- read.csv("us.csv")

# load us-states.csv data
states <- read.csv("us-states.csv")
```

The original `us` dataset contains `359` observations and `3` variables. These variables include `date` (daily dates recorded from 01/21/2020 to 01/13/2021), `cases` (cumulative cases), and `deaths` (cumulative deaths). There were no missing (NA's) values in the dataset. 

The original `states` dataset contains `17449` observations and `5` variables. These variables include `date` (daily dates recorded from 01/21/2020 to 01/13/2021), `state` (state names), `fips` (standard geographic identifier), `cases` (cumulative cases), and `deaths` (cumulative deaths). Similarly to the `us` dataset, the `states` dataset didn't have any missing (NA's) values. 

```{r, warning = FALSE}
# summary statistics of original "us" data
dim(us)
str(us)
any(is.na(us))

# summary statistics of "states" data
dim(states)
str(states)
any(is.na(states))
```

## Data Cleaning

### Clean `us` Data

```{r, warning = FALSE}
# set "date" column to standardized date format
us$date <- mdy(us$date)

# create a new data frame that includes new columns and filters "2020" as the year for data analysis
us1 <- us %>% 
  mutate(month = floor_date(date, "month"), 
         chr_month = format(date, "%B"), 
         year = format(date, "%Y")) %>% 
  filter(year == "2020")

head(us1)
```

The cleaned and filtered national-level data has been renamed to `us1`. For the `cases` column, the mean is `4992201`, the median is `3289376`, and the standard deviation is `5174886`. For the `deaths` column, the mean is `134142.8`, the median is `134779.5`, and the standard deviation is`99246.17`.

To view the full summary statistics of the data (i.e., mean, median, etc),  `summary(us1)` would be the code used to generate the results for all columns in the dataset. 

To view each individual value, `mean(us1$cases)` would be used to compute the mean of cases in a R code chunk, `median(us1$cases)` to compute the median of cases, `sd(us1$cases)` to compute the standard deviation of cases, `mean(us1$deaths)` to compute the mean of deaths, `median(us1$deaths)` to compute the median of deaths, and `sd(us1$deaths)` to compute the standard deviation of deaths. 

```{r, echo = TRUE, results = 'hide'}
# summary statistics of cleaned and filtered "us1" data
summary(us1)
any(is.na(us1))
```

```{r}
mean(us1$cases) # mean of cases
median(us1$cases) # median of cases
sd(us1$cases) # sd of cases

mean(us1$deaths) # mean of deaths
median(us1$deaths) # median of deaths
sd(us1$deaths) # sd of deaths
```

### Clean `states` Data

```{r, warning = FALSE}
#set "date" column to standardized date format
states$date <- as.Date(states$date, format = "%Y-%m-%d")

# create a new data frame that includes new columns and filters "2020" as the year for data analysis
states1 <- states %>% 
  mutate(month = floor_date(date, "month"), 
         chr_month = format(date, "%B"), 
         year = format(date, "%Y")) %>% 
  filter(year == "2020")

head(states1)
```

The cleaned and filtered state-level data has been renamed to `states1`. For the `cases` column, the mean is `103221.1`, the median is `29701`, and the standard deviation is `191944.1`. For the `deaths` column, the mean is `2773.6`, the median is `693`, and the standard deviation is`5280.142`. There are no descriptive statistics for the `date`, `month`, `chr_month`, and `year` columns since they either a date or a character column. In addition, descriptive statistics will not be computed for `fips` as it is a geographic identifieer. 

To view the full summary statistics of the data (i.e., mean, median, quantiles etc), `summary(us1)` would be the code used to generate the results for all columns in the dataset. 

To view each individual value, `mean(states1$cases)` would be used to compute the mean of cases in a R code chunk, `median(states1$cases)` to compute the median of cases, `sd(states1$cases)` to compute the standard deviation of cases, `mean(states1$deaths)` to compute the mean of deaths, `median(states1$deaths)` to compute the median of deaths, and `sd(states1$deaths)` to compute the standard deviation of deaths. 

```{r, echo = TRUE, results = 'hide'}
# summary statistics of cleaned and filtered "states1" data
summary(states1)
any(is.na(states1))
```

```{r}
mean(states1$cases) # mean of cases
median(states1$cases) # median of cases
sd(states1$cases) # sd of cases

mean(states1$deaths) # mean of deaths
median(states1$deaths) # median of deaths
sd(states1$deaths) # sd of deaths
```

# Exploratory Data Analysis

The goal of this project is to investigate the impact of the COVID-19 pandemic on United States and New York by assessing the 2020 monthly percent increase in cases and deaths. We will not be performing any data analysis on 2021 due the fact that we do not have enough information as data is only available until January 13, 2021. 

## US % Change in COVID-19 Cases Computation

```{r, warning = FALSE}
us_monthly_cases <- us1 %>% 
  group_by(month) %>% 
  summarize(begin_cases = first(cases)) %>% 
  ungroup()

us_monthly_cases <- us_monthly_cases %>% 
  arrange(month) %>% 
  mutate(new_cases = lead(begin_cases, default = last(us$cases)) - begin_cases)

us_monthly_cases <- us_monthly_cases %>%
  arrange(month) %>%
  mutate(
    prev_new_cases = lag(new_cases, default = new_cases[1]), 
    percent_increase = (new_cases / prev_new_cases) * 100
  ) %>%
  mutate(
    percent_increase = ifelse(is.infinite(percent_increase) | is.nan(percent_increase), 
                              NA, percent_increase)
  ) 
```

## New York % Change in COVID-19 Cases Computation

```{r, warning = FALSE}
states_monthly_cases <- states1 %>% 
  group_by(month) %>%
  filter(state == "New York") %>% 
  summarize(begin_cases = first(cases)) %>% 
  ungroup()

states_monthly_cases <- states_monthly_cases %>% 
  arrange(month) %>% 
  mutate(new_cases = lead(begin_cases, default = last(states$cases)) - begin_cases)

states_monthly_cases <- states_monthly_cases %>%
  arrange(month) %>%
  mutate(
    prev_new_cases = lag(new_cases, default = new_cases[1]), 
    percent_increase = (new_cases / prev_new_cases) * 100
  ) %>%
  mutate(
    percent_increase = ifelse(is.infinite(percent_increase) | is.nan(percent_increase), 
                              NA, percent_increase)
  )
```

## US and New York % Change in COVID-19 Cases Visualizations

```{r, warning = FALSE}
us_monthly_cases$month <- as.Date(us_monthly_cases$month)
us_monthly_cases$month_name <- factor(format(us_monthly_cases$month, "%B"), 
                                      levels = month.name)

states_monthly_cases$month <- as.Date(states_monthly_cases$month)
states_monthly_cases$month_name <- factor(format(states_monthly_cases$month, "%B"), 
                                          levels = month.name)

# table
us_monthly_cases_1 <- us_monthly_cases %>% 
  select(month_name, percent_increase)

states_monthly_cases_1 <- states_monthly_cases %>% 
  select(month_name, percent_increase) 

combined_cases <- merge(us_monthly_cases_1, states_monthly_cases_1, by = "month_name", all = TRUE)

names(combined_cases)[names(combined_cases) == "month_name"] <- "Month"
names(combined_cases)[names(combined_cases) == "percent_increase.x"] <- "US"
names(combined_cases)[names(combined_cases) == "percent_increase.y"] <- "New York"

knitr::kable(combined_cases, caption = "% Change in COVID-19 Cases for US and New York", align = c("c", "c", "c"))

# graph
us_monthly_cases$Type <- 'US'
states_monthly_cases$Type <- 'New York'

combined_data <- rbind(us_monthly_cases, states_monthly_cases)

ggplot(combined_data, aes(x = month_name, y = percent_increase, fill = Type)) +
  geom_bar(stat = "identity", position = "dodge") +
  theme_minimal() +
  labs(title = "Figure 1: US vs. New York Monthly % Change in COVID-19 Cases", 
       x = "Month", y = "Percent Change (%)") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "bottom") +
  scale_fill_manual(values = c("US" = "orange", "New York" = "steelblue")) +
  ylim(0, 1200)
```

US experienced an exceptionally high % increase in cases in the early months of the pandemic, especially noticeable in February and March 2020. However, the % increase remained low from May to October compared to the other months. New York, on the other hand, experienced the sudden % increase in cases once in April and towards the last couple of months, especially in October and November 2020. 

The variability in the data could be attributed to various factors, such as public health interventions (i.e., social distancing, the use of face masks, etc.) or changes in data reporting (i.e., over-reporting, under-reporting, etc.). 

There are some extreme values (both positive and negative) and missing (NA's) values which will need to be addressed in further analyses to improve on the data presented. 

---

## US % Change in COVID-19 Deaths Computation

```{r, warning = FALSE}
us_monthly_deaths <- us1 %>% 
  group_by(month) %>% 
  summarize(begin_deaths = first(deaths)) %>% 
  ungroup()

us_monthly_deaths <- us_monthly_deaths %>% 
  arrange(month) %>% 
  mutate(new_deaths = lead(begin_deaths, default = last(us$deaths)) - begin_deaths)

us_monthly_deaths <- us_monthly_deaths %>%
  arrange(month) %>%
  mutate(
    prev_new_deaths = lag(new_deaths, default = new_deaths[1]), 
    percent_increase = (new_deaths / prev_new_deaths) * 100
  ) %>%
  mutate(
    percent_increase = ifelse(is.infinite(percent_increase) | is.nan(percent_increase), 
                              NA, percent_increase)
  ) 
```

## New York % Change in COVID-19 Deaths Computation

```{r, warning = FALSE}
states_monthly_deaths <- states1 %>% 
  group_by(month) %>%
  filter(state == "New York") %>% 
  summarize(begin_deaths = first(deaths)) %>% 
  ungroup()

states_monthly_deaths <- states_monthly_deaths %>% 
  arrange(month) %>% 
  mutate(new_deaths = lead(begin_deaths, default = last(states$deaths)) - begin_deaths)

states_monthly_deaths <- states_monthly_deaths %>%
  arrange(month) %>%
  mutate(
    prev_new_deaths = lag(new_deaths, default = new_deaths[1]), 
    percent_increase = (new_deaths / prev_new_deaths) * 100
  ) %>%
  mutate(
    percent_increase = ifelse(is.infinite(percent_increase) | is.nan(percent_increase), 
                              NA, percent_increase)
  )
```

## US vs. New York % Change in COVID-19 Deaths Visualizations

```{r, warning=FALSE}
us_monthly_deaths$month <- as.Date(us_monthly_deaths$month)
us_monthly_deaths$month_name <- factor(format(us_monthly_deaths$month, "%B"),
                                       levels = month.name)

states_monthly_deaths$month <- as.Date(states_monthly_deaths$month)
states_monthly_deaths$month_name <- factor(format(states_monthly_deaths$month, "%B"), 
                                           levels = month.name)

# table
us_monthly_deaths_1 <- us_monthly_deaths %>% 
  select(month_name, percent_increase)

states_monthly_deaths_1 <- states_monthly_deaths %>% 
  select(month_name, percent_increase) 

combined_deaths <- merge(us_monthly_deaths_1, states_monthly_deaths_1, by = "month_name", all = TRUE)

names(combined_deaths)[names(combined_deaths) == "month_name"] <- "Month"
names(combined_deaths)[names(combined_deaths) == "percent_increase.x"] <- "US"
names(combined_deaths)[names(combined_deaths) == "percent_increase.y"] <- "New York"

knitr::kable(combined_deaths, caption = "% Change in COVID-19 Deaths for US and New York", align = c("c", "c", "c"))

# graph
us_monthly_deaths$Type <- 'US'
states_monthly_deaths$Type <- 'New York'

combined_data <- rbind(us_monthly_deaths, states_monthly_deaths)

ggplot(combined_data, aes(x = month_name, y = percent_increase, fill = Type)) +
  geom_bar(stat = "identity", position = "dodge") +
  theme_minimal() +
  labs(title = "Figure 2: US vs. New York Monthly % Change in COVID-19 Deaths", 
       x = "Month", y = "Percent Change (%)") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "bottom") +
  scale_fill_manual(values = c("US" = "orange", "New York" = "steelblue")) +
  ylim(0, 1200)
```

US experienced an exceptionally high % increase in deaths in the early months of the pandemic, especially noticeable in March and April 2020. However, the % increase remained low from May to October compared to the other months. New York experienced high % increase in deaths in April and November 2020. 

The variability in the data could be attributed to various factors, such as
rapid improvements in healthcare to address the COVID-19 pandemic (i.e., vaccinations, medications, etc.) or changes in data reporting (i.e., over-reporting, under-reporting, etc.). 

There are some extreme values (both positive and negative) and missing (NA's) values which will need to be addressed in further analyses to improve on the data presented. 

# Summary

The outbreak of the COVID-19 pandemic was detected from the steep % increases in early 2020 for the US and New York, indicative of the virus's rapid spread. The US recorded its peak increases in February and March, while New York's numbers soared in April and fluctuated thereafter. The data illustrated New York's situation improved relative to the rest of the US, where variability in case changes persisted, suggesting the influence of public health measurees and the pandemic's evolution. 

However, the analysis is limited by potential inconsistencies in testing and reporting, presented by having missing (NA's) and extreme (both positive and negative) values. It's also lacking other variables such as population density or public health policies. Future research should incorporate these factors as well as extending the timeframe beyond 2020 to have a more comprehensive analysis. 